---
title: "Batch Correction"
output:
  html_document:
    df_print: paged
---

This notebook serves to compute batch corrected read counts from nf-core/rnaseq/featureCounts output.
The raw read counts will be batch corrected and then converted to TPM for clustering analysis and 
ML in python.


```{r Libraries}
library(tidyverse)
library(sva)
library(tibble)
```


# Read FeatureCounts Data

```{r Reading data}

pancreas_data <- read.table("/Users/susanne/Documents/Master/Data/pancreas/merged_featureCounts_all_pancreas.txt", sep="\t", header=TRUE)
pheno <- read.table("/Users/susanne/Documents/Master/Data/pancreas/metadata_all_pancreas_with_batches.csv", sep=";", header=TRUE)
gene_lengths <- read.table("/Users/susanne/Documents/Master/Data/pancreas/featureCounts_geneLengths.txt", sep= ",", header=TRUE)

# tidiverse object better for selecting data
pancreas<- as_tibble(pancreas_data)
# gene_ids, gene_names
gene_id <- pancreas$GeneID
gene_name <- pancreas$GeneName

# selection of samples without gene_id, gene_name
expr_df <- pancreas %>% select(3:ncol(pancreas))

# ADJUST SAMPLE IDS BACK TO NORMAL
# replace X in sample_name strings by ""
sample_names <- sub("X", "", colnames(expr_df))
# replace . by - in sample ids
sample_names <- sample_names %>% str_replace_all("\\.", "\\-") 


# sort sample names of pheno_table with batches according to order in pancreas_data‚
sorted_pheno <- pheno[order(match(pheno[,1],sample_names)),]
# sort geneIds of featureCounts_geneLenghts according to pancreas data GeneIds
sorted_gene_lengths <- gene_lengths[order(match(gene_lengths[,1], pancreas_data$GeneID)),]


# vector of batches
batch_pancreas = sorted_pheno$Batch

```


##Function

* [source of TPM function](https://gist.github.com/slowkow/c6ab0348747f86e2748b)
* [source normalization with min max ](https://stackoverflow.com/questions/44050028/min-max-scaling-normalization-in-r-for-train-and-test-data)

```{r Functions}

# TPM Function from 
  tpm <- function(counts, lengths) {
  rate <- counts / lengths
  rate / sum(rate) * 1e6
  }

# mimics scikit-learn.preprocessing.MinMaxScaler()
# mi, ma = range(mi, ma)
normalize <- function(x, mi, ma)
{
    X_sc <- (x- min(x)) / (max(x)-min(x))
    X_sc * (ma-mi) + mi
}

```



# SVAseq: 

## ComBat 

```{r ComBat}

# parametric adjustment, consider plotting plots
combat_edata1 = ComBat(dat=expr_df, batch=batch_pancreas, mod=NULL, par.prior=TRUE, prior.plots=FALSE)

# convert combat batch corrected values to tpm
combat_tpm_pancreas <- apply(combat_edata1, 2, tpm, lengths=sorted_gene_lengths$GeneLength)
# convert to dataframe
combat_pancreas_tpm_df <- data.frame(combat_tpm_pancreas)
combat_df <- data.frame(combat_tpm_pancreas)
# append GeneId, GeneName to sample_names
table_header <- append(c("GeneID", "GeneName"), sample_names, after=2)

# add columns GeneId, GeneName to dataframe 
combat_pancreas_tpm_df <- add_column(combat_pancreas_tpm_df, gene_id, .before=1)
combat_pancreas_tpm_df <- add_column(combat_pancreas_tpm_df, gene_name, .before=2)

colnames(combat_pancreas_tpm_df) <- table_header
# write.table(combat_df, "/Users/susanne/Documents/Master/Data/pancreas/ComBat_corrected_TPM_all_pancreas.txt", sep="\t", row.names = FALSE, col.names =  TRUE, quote=FALSE)



```


## ComBat with model

```{r ComBat model}
# intercept, cancer, normal
mod = model.matrix(~as.factor(Sample_Type), data=sorted_pheno)

combat_edata3 = ComBat(dat=expr_df, batch=batch_pancreas, mod=mod, par.prior=TRUE)

# convert combat batch corrected values to tpm
combat_tpm_pancreas3 <- apply(combat_edata3, 2, tpm, lengths=sorted_gene_lengths$GeneLength)

# convert to dataframe
combat_pancreas_tpm_df3 <- data.frame(combat_tpm_pancreas3)
combat_model <- data.frame(combat_tpm_pancreas3)
# append GeneId, GeneName to sample_names
table_header <- append(c("GeneID", "GeneName"), sample_names, after=2)

# add columns GeneId, GeneName to dataframe 
combat_pancreas_tpm_df3 <- add_column(combat_pancreas_tpm_df3, gene_id, .before=1)
combat_pancreas_tpm_df3 <- add_column(combat_pancreas_tpm_df3, gene_name, .before=2)

colnames(combat_pancreas_tpm_df3) <- table_header
# write.table(combat_pancreas_tpm_df3, "/Users/susanne/Documents/Master/Data/pancreas/ComBatModel_corrected_TPM_all_pancreas.txt", sep="\t", row.names = FALSE, col.names =  TRUE, quote=FALSE)

```


## CombatSeq

```{r CombatSeq}

# combat-seq takes as input a count matrix
expr_mat <- as.matrix(expr_df)

# use sorted batch variable
condition <- as.numeric(as.factor(sorted_pheno$Sample_Type))

# perform batch correction with condition and batches specified
combatseq_counts_pancreas <- ComBat_seq(expr_mat, batch=batch_pancreas, group=condition, full_mod=TRUE)

# conversion to TPM
# APPLY(DATA, MARGIN (1 = ROWS, 2 = COLUMNS), FUNCTION)
combatseq_tpm_pancreas <- apply(combatseq_counts_pancreas, 2, tpm, lengths=sorted_gene_lengths$GeneLength)

# convert to dataframe
combatseq_pancreas_tpm_df <- data.frame(combatseq_tpm_pancreas)
combatseq_df <- data.frame(combatseq_tpm_pancreas)


# add columns GeneId, GeneName to dataframe 
combatseq_pancreas_tpm_df <- add_column(combatseq_pancreas_tpm_df, gene_id, .before=1)
combatseq_pancreas_tpm_df <- add_column(combatseq_pancreas_tpm_df, gene_name, .before=2)

colnames(combatseq_pancreas_tpm_df) <- table_header

# write.table(combatseq_df, "/Users/susanne/Documents/Master/Data/pancreas/CombatSeq_corrected_TPM_all_pancreas.txt", sep="\t", row.names = FALSE, col.names =  TRUE, quote=FALSE)


```



## PCA analysis prcomp

```{r prcomp}

library(ggplot2)
library(grid)
library(gridExtra)
library(ggpubr)
library(data.table)

# -----------------------------------------------------------------------------
# 1) uncorrected TPM values

# convert counts to TPM 
expr_TPM <- apply(expr_df, 2, tpm, lengths=sorted_gene_lengths$GeneLength)


# convert to dataframe
expr_df_TPM <- data.frame(expr_TPM)


# write exprTPM to output
# convert to dataframe
uncorr_TPM_df <- data.frame(expr_TPM)
# add columns GeneId, GeneName to dataframe 
uncorr_TPM_df <- add_column(uncorr_TPM_df, gene_id, .before=1)
uncorr_TPM_df <- add_column(uncorr_TPM_df, gene_name, .before=2)
colnames(uncorr_TPM_df) <- table_header
# write.table(uncorr_TPM_df, "/Users/susanne/Documents/Master/Data/pancreas/Uncorrected_TPM_all_pancreas_fCounts.txt", sep="\t", row.names = FALSE, col.names =  TRUE, quote=FALSE)


expr_df_TPM_t <- transpose(expr_df_TPM)

# attach sample names and feature names
row.names(expr_df_TPM_t) <- sample_names
colnames(expr_df_TPM_t) <- gene_id



# get condition
sample_condition <- sorted_pheno$Sample_Type

# PCA uncorrected data
df_pca_uncorr <- prcomp(expr_df_TPM_t)

# plot PCA per condition
df_out_uncorr <- as.data.frame(df_pca_uncorr$x)
df_out_uncorr$condition <- sample_condition
pca_uncorr_c<-ggplot(df_out_uncorr,aes(x=PC1,y=PC2,color=condition ))
pca_uncorr_c<-pca_uncorr_c+geom_point()

# PCA per project
df_out_uncorr <- as.data.frame(df_pca_uncorr$x)
df_out_uncorr$project <- sorted_pheno$Project
pca_uncorr_p<-ggplot(df_out_uncorr,aes(x=PC1,y=PC2,color=project ))
pca_uncorr_p<-pca_uncorr_p+geom_point()


# -----------------------------------------------------------------------------
# 2) Combat corrected TPM 
combat_pancreas_tpm_df_t <- transpose(combat_df)
row.names(combat_pancreas_tpm_df_t) <- sample_names
colnames(combat_pancreas_tpm_df_t) <- gene_id

df_pca_combat <- prcomp(combat_pancreas_tpm_df_t)

# PCA combat per condition
df_out_combat <- as.data.frame(df_pca_combat$x)
df_out_combat$condition <- sample_condition
pca_combat_c<-ggplot(df_out_combat,aes(x=PC1,y=PC2,color=condition ))
pca_combat_c<-pca_combat_c+geom_point()

# PCA combat per project
df_out_combat <- as.data.frame(df_pca_combat$x)
df_out_combat$project <- sorted_pheno$Project
pca_combat_p<-ggplot(df_out_combat,aes(x=PC1,y=PC2,color=project ))
pca_combat_p<-pca_combat_p+geom_point()

# -----------------------------------------------------------------------------
# 3) Combat corrected with model TPM
combat_model_pancreas_tpm_t <- transpose(combat_model)
row.names(combat_model_pancreas_tpm_t) <- sample_names
colnames(combat_model_pancreas_tpm_t) <- gene_id

# PCA
df_pca_combat_model <- prcomp(combat_model_pancreas_tpm_t)

# PCA combat model per condition
df_out_combat_model <- as.data.frame(df_pca_combat_model$x)
df_out_combat_model$condition <- sample_condition
pca_combat_model_c<-ggplot(df_out_combat_model,aes(x=PC1,y=PC2,color=condition ))
pca_combat_model_c<-pca_combat_model_c+geom_point()

# PCA combat model per project
df_out_combat_model <- as.data.frame(df_pca_combat_model$x)
df_out_combat_model$project <- sorted_pheno$Project
pca_combat_model_p<-ggplot(df_out_combat_model,aes(x=PC1,y=PC2,color=project ))
pca_combat_model_p<-pca_combat_model_p+geom_point()

# -----------------------------------------------------------------------------
# 4) CombatSeq corrected TPM

combatseq_t <- transpose(combatseq_df)
row.names(combatseq_t) <- sample_names
colnames(combatseq_t) <- gene_id

# PCA
df_pca_combatseq <- prcomp(combatseq_t)


# PCA combatseq condition
df_out_combatseq <- as.data.frame(df_pca_combatseq$x)
df_out_combatseq$condition <- sample_condition
pca_combatseq_c<-ggplot(df_out_combatseq,aes(x=PC1,y=PC2,color=condition ))
pca_combatseq_c<-pca_combatseq_c+geom_point()

# PCA combatseq project
df_out_combatseq <- as.data.frame(df_pca_combatseq$x)
df_out_combatseq$project <- sorted_pheno$Project
pca_combatseq_p<-ggplot(df_out_combatseq,aes(x=PC1,y=PC2,color=project ))
pca_combatseq_p<-pca_combatseq_p+geom_point()



# draw multipage
figure1 <- ggarrange(pca_uncorr_c, pca_uncorr_p,
          pca_combat_c, pca_combat_p,
          pca_combat_model_c, pca_combat_model_p,
          pca_combatseq_c, pca_combatseq_p,
          labels = c("U", "U", 
                     "C", "C",
                     "CM", "CM",
                     "CS", "CS"),
                        nrow = 4, ncol = 2)

figure1 <- annotate_figure(figure1, top = text_grob("PCA of unscaled data", color = "black", face = "bold", size = 12))
ggexport(figure1, filename = "/Users/susanne/Downloads/pca_unscaled_condition_project.png")
```


## Scaling to MinMax

```{r scaling}
#               UNCORRECTED DATA
# -----------------------------------------------------------------------------

df_uncorr_scaled <- apply(expr_df_TPM_t, 2, normalize, -1,1)

# replace infinity and is.na with 0
is.na(df_uncorr_scaled) <- sapply(df_uncorr_scaled, is.infinite)
df_uncorr_scaled[is.na(df_uncorr_scaled)] <- 0

# PCA uncorrected data
df_pca_uncorr_scaled <- prcomp(df_uncorr_scaled)

# plot PCA per condition
df_out_uncorr_scaled <- as.data.frame(df_pca_uncorr_scaled$x)
df_out_uncorr_scaled$condition <- sample_condition
pca_uncorr_scaled_c<-ggplot(df_out_uncorr_scaled,aes(x=PC1,y=PC2,color=condition ))
pca_uncorr_scaled_c<-pca_uncorr_scaled_c+geom_point() 

# PCA per project
df_out_uncorr_scaled <- as.data.frame(df_pca_uncorr_scaled$x)
df_out_uncorr_scaled$project <- sorted_pheno$Project
pca_uncorr_scaled_p<-ggplot(df_out_uncorr_scaled,aes(x=PC1,y=PC2,color=project ))
pca_uncorr_scaled_p<-pca_uncorr_scaled_p+geom_point()


#               COMBAT CORRECTED AND SCALED
# -----------------------------------------------------------------------------


df_combat_scaled <- apply(combat_pancreas_tpm_df_t, 2, normalize, -1,1)

# replace infinity and is.na with 0
is.na(df_combat_scaled) <- sapply(df_combat_scaled, is.infinite)
df_combat_scaled[is.na(df_combat_scaled)] <- 0

# PCA combat scaled
pca_combat_sc <- prcomp(df_combat_scaled)

# PCA combat per condition
df_out_combat_sc <- as.data.frame(pca_combat_sc$x)
df_out_combat_sc$condition <- sample_condition
pca_combat_sc_c<-ggplot(df_out_combat_sc,aes(x=PC1,y=PC2,color=condition ))
pca_combat_sc_c<-pca_combat_sc_c+geom_point()

# PCA combat per project

df_out_combat_sc <- as.data.frame(pca_combat_sc$x)
df_out_combat_sc$project <- sorted_pheno$Project
pca_combat_sc_p<-ggplot(df_out_combat_sc,aes(x=PC1,y=PC2,color=project ))
pca_combat_sc_p<-pca_combat_sc_p+geom_point()




#               COMBAT WITH MODEL CORRECTED AND SCALED
# -----------------------------------------------------------------------------
df_combat_model_scaled <- apply(combat_model_pancreas_tpm_t, 2, normalize, -1,1)

is.na(df_combat_model_scaled) <- sapply(df_combat_model_scaled, is.infinite)
df_combat_model_scaled[is.na(df_combat_model_scaled)] <- 0


# PCA combat model scaled
pca_combat_model_sc <- prcomp(df_combat_model_scaled)

# PCA combat per condition
df_out_combat_model_sc <- as.data.frame(pca_combat_model_sc$x)
df_out_combat_model_sc$condition <- sample_condition
pca_combat_model_sc_c<-ggplot(df_out_combat_model_sc,aes(x=PC1,y=PC2,color=condition ))
pca_combat_model_sc_c<-pca_combat_model_sc_c+geom_point()

# PCA combat per project

df_out_combat_model_sc <- as.data.frame(pca_combat_model_sc$x)
df_out_combat_model_sc$project <- sorted_pheno$Project
pca_combat_model_sc_p<-ggplot(df_out_combat_model_sc,aes(x=PC1,y=PC2,color=project ))
pca_combat_model_sc_p<-pca_combat_model_sc_p+geom_point()


#               COMBATSEQ CORRECTED AND SCALED
# -----------------------------------------------------------------------------


combatseq_t_sc <- apply(combatseq_t, 2, normalize, -1,1)

is.na(combatseq_t_sc) <- sapply(combatseq_t_sc, is.infinite)
combatseq_t_sc[is.na(combatseq_t_sc)] <- 0

# PCA combatseq scaled
pca_combatseq_sc <- prcomp(combatseq_t_sc)

# PCA combatseq per condition
df_combatseq_sc <- as.data.frame(pca_combatseq_sc$x)
df_combatseq_sc$condition <- sample_condition
pca_combatseq_sc_c<-ggplot(df_combatseq_sc,aes(x=PC1,y=PC2,color=condition ))
pca_combatseq_sc_c<-pca_combatseq_sc_c+geom_point()

# PCA combatseq per project

df_combatseq_sc <- as.data.frame(pca_combatseq_sc$x)
df_combatseq_sc$project <- sorted_pheno$Project
pca_combatseq_sc_p<-ggplot(df_combatseq_sc,aes(x=PC1,y=PC2,color=project ))
pca_combatseq_sc_p<-pca_combatseq_sc_p+geom_point()
pca_combatseq_sc_p + ggtitle("CombatSeq") 



# ----------------------------------------------------------------------------
# draw multipage
figure2 <- ggarrange(pca_uncorr_scaled_c, pca_uncorr_scaled_p,
          pca_combat_sc_c, pca_combat_sc_p,
          pca_combat_model_sc_c, pca_combat_model_sc_p,
          pca_combatseq_sc_c, pca_combatseq_sc_p,
          labels = c("U", "U", 
                     "C", "C",
                     "CM", "CM",
                     "CS", "CS"),
                        nrow = 4, ncol = 2)

figure2 <- annotate_figure(figure2, top = text_grob("PCA of MinMax-scaled data", color = "black", face = "bold", size = 12))

ggexport(figure2, filename = "/Users/susanne/Downloads/pca_scaled_condition_project.png")

```

